---
title: "Baseball Analytics 1"
author: "Mendoza College of Business"
date: "10/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's first load the packages we will use for this analysis:

```{r}
library(ggplot2) # Load ggplot2
library(reshape) # Load tidyverse
```

# Set Up

For this analysis we will work with statcast data. Statcast is the tracking technology that is used to collect baseball data for the MLB. The data can be downloaded into R using the `baseballr` package. The script used to download and prepare the data for this class is `baseball_data_scrapr.r`.

There are a few pre-processing steps to go through which we can skip for the class and load in the modified data stored as `statcast_2022.rda`. This dataset contains all of the pitches from the 2021 MLB season. 

Let's load it in:

```{r}
load("2022_stat_cast_data.rda") # Load data
```


Load in all batter data for the Chicago Cubs


```{r}
cubs <- batters[batters$batter_team == "CHC", ]
```

Series I have chosen is a series from July 12-13, 2022 against the Baltimore Orioles

July 12,

Winning pitcher was Lyles, C. Perez with a hold, and Jorge Lopez with the save

July 13, 




```{r}

table(cubs$type)

single <- cubs[cubs$events == "single",]
double <- cubs[cubs$events == "double",]
triple <- cubs[cubs$events == "triple",]
HR <- cubs[cubs$events == "home_run",]

singleL <- single[single$p_throws == "L",]
SingleR <- single[single$p_throws == "R",]
doubleL <- double[double$p_throws == "L",]
doubleR <- double[double$p_throws == "R",]
tripleL <- triple[triple$p_throws == "L",]
tripleR <- triple[triple$p_throws == "R",]
HRL <- HR[HR$p_throws == "L",]
HRR <- HR[HR$p_throws == "R",]

HitR <- 13 + 400 + 66 + 118
HitL <- 7 + 132 + 31 + 49


```


Looking at the spread of hits it seems that they hit much better off Right handed pitchers, but that could be attributed to the fact they face a lot more Right handed pitchers, let's look.


```{r}
pitchL <- cubs[cubs$p_throws == "L",]
pitchR <- cubs[cubs$p_throws == "R",]

teamavgL <- .219
teamavgR <- .231
```

I was correct, the cubs face a lot more right handed pitches than they do left handed pitches. My next step is I want to convert these to expected bases per pitch for both left and right handers. 


```{r}
leftxbp <- (nrow(singleL)/nrow(pitchL)) + ((nrow(doubleL)/nrow(pitchL) ) * 2) + ((nrow(tripleL)/nrow(pitchL) ) * 3) + ((nrow(HRL)/nrow(pitchL) ) * 4)


rightxbp <- (nrow(SingleR)/nrow(pitchR)) + ((nrow(doubleR)/nrow(pitchR) ) * 2) + ((nrow(tripleR)/nrow(pitchR) ) * 3) + ((nrow(HRR)/nrow(pitchR) ) * 4)


leftxbp
rightxbp
```

After running this we see that in fact the cubs are expected to get more bases against left handed pitchers compared to right handed, this could be a contributing factor in why they see more right handed pitchers. Next I want to focus on the type of pitches the players hit the best


```{r}

library(stringr)

hit <- cubs[cubs$events == "single" | cubs$events == "double" | cubs$events == "triple" | cubs$events == "home_run", ]

```



Time to focus on the pitchers from the selected games

July 12 
Winning pitcher was Lyles, C. Perez with a hold, and Jorge Lopez with the save


```{r}
Lyles <- pitchers[pitchers$player_name == "Lyles, Jordan",]
XBaLyles <- (teamavgR+.279)/2

Perez <- pitchers[pitchers$pitcher == 672335,]
Lopez <- pitchers[pitchers$pitcher == 605347,]

hitLyles <- Lyles[Lyles$events == "single" | Lyles$events == "double" | Lyles$events == "triple" | Lyles$events == "home_run",]

hitPerez <- Perez[Perez$events == "single" | Perez$events == "double" | Perez$events == "triple" | Perez$events == "home_run",]

hitLopez <- Lopez[Lopez$events == "single" | Lopez$events == "double" | Lopez$events == "triple" | Lopez$events == "home_run",]

```

July 13

Watkins, Tate, C. Perez, Krehbiel, Baker

```{r}
Watkins <- pitchers[pitchers$player_name == "Watkins, Spenser",]
Tate <- pitchers[pitchers$player_name == "Tate, Dillon",]
Krehbiel <- pitchers[pitchers$player_name == "Krehbiel, Joey",]
Baker <- pitchers[pitchers$player_name == "Baker, Bryan",]


hitWatkins <- Watkins[Watkins$events == "single" | Watkins$events == "double" | Watkins$events == "triple" | Watkins$events == "home_run",]

hitTate <- Tate[Tate$events == "single" | Tate$events == "double" | Tate$events == "triple" | Tate$events == "home_run",]

hitKrehbiel <- Krehbiel[Krehbiel$events == "single" | Krehbiel$events == "double" | Krehbiel$events == "triple" | Krehbiel$events == "home_run",]

hitBaker <- Baker[Baker$events == "single" | Baker$events == "double" | Baker$events == "triple" | Baker$events == "home_run",]
```


Now I want to do something similar for pitchers as I did batters, I want to see the expected hits per pitch


```{r}
Lylesxbp <- (nrow(hitLyles)/nrow(Lyles))
Perezxbp <- (nrow(hitPerez)/nrow(Perez))
Lopezxbp <- (nrow(hitLopez)/nrow(Lopez))
Watkinsxbp <- (nrow(hitWatkins)/nrow(Watkins))
Tatexbp <- (nrow(hitTate)/nrow(Tate))
Krehbielxbp <- (nrow(hitKrehbiel)/nrow(Krehbiel))
Bakerxbp <- (nrow(hitBaker)/nrow(Baker))

```

From looking at this Lopez has the lowest expected hits per pitch, this is expected as he is a closing pitcher.

Calculating the orioles batting averages vs right handed pitchers along with other hitting stats

```{r}
H <- 92 + 105 +98+119+84+74+72+65+79+27+16+22+9+5+21+0+4+0+2+3+20+16+1+2+0

PA <- 417+413+389+426+300+286+342+235+346+93+106+89+29+27+80+6+41+1+10+21+93+92+27+16+3

OBa <- H/PA

XBaSampson <- (.253+OBa)/2
XBaEffross <- (.220+OBa)/2
XBaGivens <- (.216+OBa)/2
XBaSwarmer <- (.250+OBa)/2

orioles <- batters[batters$batter_team == "BAL", ]

singleO <- orioles[orioles$events == "single",]
doubleO <- orioles[orioles$events == "double",]
tripleO <- orioles[orioles$events == "triple",]
HRO <- orioles[orioles$events == "home_run",]

singleLO <- singleO[singleO$p_throws == "L",]
SingleRO <- singleO[singleO$p_throws == "R",]
doubleLO <- doubleO[doubleO$p_throws == "L",]
doubleRO <- doubleO[doubleO$p_throws == "R",]
tripleLO <- tripleO[tripleO$p_throws == "L",]
tripleRO <- tripleO[tripleO$p_throws == "R",]
HRLO <- HRO[HRO$p_throws == "L",]
HRRO <- HRO[HRO$p_throws == "R",]

Hit0 <- 8 +313+130+70



```

Cubs vs left handers because Cionel Perez is a lefty, then followed by Jorge Lopez


```{r}
HLC <- 31+23+20+28+14+19+16+39+37+10+16+1+6+1+5+20+3+4+3+2+1+7
PALC <- 124+105+83+104+71+100+76+128+126+32+73+8+33+5+24+27+23+78+14+19+6+29+20+35

CLBa <- HLC/PALC

xBaPerez <- (.220+CLBa)/2


XBaLopez <- (teamavgR+.174)/2




```


I am going to try and start a Monte Carlo simulation here, not about to be very easy.



```{r}
set.seed(09161999)

finalC <- rep(NA, 5000)
finalO <- rep(NA, 5000)

scores.df <- data.frame(finalC, finalO)

for (i in 1:nrow(scores.df)){

inningC <- 1
runsC <-  0
outsC <- 0
runnersC <- 0

while (inningC < 10){
  #Rbinom to simulate whether there was a hit or not
  hitC <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersC <- case_when(
    hitC < .171 ~ runnersC + 1,
    hitC < .221 ~ runnersC + 2,
    hitC < .227 ~ runnersC + 3,
    hitC < .256 ~ runnersC + 4,
    TRUE ~ 0)
  if (hitC >= round(XBaLyles,3)) {
    outsC <- outsC + 1
  }
  #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a runner scored
  
  if (runnersC >= 4){
    runsC = runsC + 1
    runnersC <-  runnersC - 4
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsC == 3){
    inningC = inningC + 1
    outsC <- 0
    runnersC <- 0
  }
  if (inningC == 7 | inningC == 8){
    hitC <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersC <- case_when(
    hitC < .135 ~ runnersC + 1,
    hitC < .185 ~ runnersC + 2,
    hitC < .192 ~ runnersC + 3,
    hitC < .224 ~ runnersC + 4,
    TRUE ~ 0)
  if (hitC >= round(xBaPerez,3)) {
    outsC <- outsC + 1
  }
  if (runnersC >= 4){
    runsC = runsC + 1
    runnersC <-  runnersC - 4
    
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsC == 3){
    inningC = inningC + 1
    outsC <- 0
    runnersC <- 0
  }
  
  }
  if(inningC == 9){
    hitC <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersC <- case_when(
    hitC < .136 ~ runnersC + 1,
    hitC < .176 ~ runnersC + 2,
    hitC < .180 ~ runnersC + 3,
    hitC < .203 ~ runnersC + 4,
    TRUE ~ 0)
    if (hitC >= round(XBaLopez,3)) {
    outsC <- outsC + 1
    }
  if (runnersC >= 4){
    runsC = runsC + 1
    runnersC <-  runnersC - 4
    
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsC == 3){
    inningC = inningC + 1
    outsC <- 0
    runnersC <- 0
  }
    
  }

  
}

inningO <- 1
runsO <-  0
outsO <- 0
runnersO <- 0

while (inningO < 10){
  #Rbinom to simulate whether there was a hit or not
  hitO <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersO <- case_when(
    hitO < .148 ~ runnersO + 1,
    hitO < .210 ~ runnersO + 2,
    hitO < .214 ~ runnersO + 3,
    hitO < .248 ~ runnersO + 4,
    TRUE ~ 0)
  if (hitO >= round(XBaSampson,3) ) {
    outsO <- outsO + 1
  }
  #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a runner scored
  
  if (runnersO >= 4){
    runsO = runsO + 1
    runnersO <-  runnersO - 4
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsO == 3){
    inningO = inningO + 1
    outsO <- 0
    runnersO <- 0
  }
  if(inningO == 7){
    hitO <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersO <- case_when(
    hitO < .138 ~ runnersO + 1,
    hitO < .195 ~ runnersO + 2,
    hitO < .199 ~ runnersO + 3,
    hitO < .231 ~ runnersO + 4,
    TRUE ~ 0)
  if (hitO >= round(XBaEffross,3) ) {
    outsO <- outsO + 1
  }
  #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a runner scored
  if (runnersO >= 4){
    runsO = runsO + 1
    runnersO <-  runnersO - 4
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsO == 3){
    inningO = inningO + 1
    outsO <- 0
    runnersO <- 0
  }
    
    
  }
  if(inningO == 8){
    hitO <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersO <- case_when(
    hitO < .137 ~ runnersO + 1,
    hitO < .194 ~ runnersO + 2,
    hitO < .198 ~ runnersO + 3,
    hitO < .229 ~ runnersO + 4,
    TRUE ~ 0)
  if (hitO >= round(XBaGivens, 3) ) {
    outsO <- outsO + 1
  }
  #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a runner scored
  if (runnersO >= 4){
    runsO = runsO + 1
    runnersO <-  runnersO - 4
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsO == 3){
    inningO = inningO + 1
    outsO <- 0
    runnersO <- 0
  }
  }
  if (inningO == 9){
    hitO <- runif(1,0,1)
  #Check if it was a hit or an out, if a hit we add one base runner
  runnersO <- case_when(
    hitO < .147 ~ runnersO + 1,
    hitO < .208 ~ runnersO + 2,
    hitO < .212 ~ runnersO + 3,
    hitO < .246 ~ runnersO + 4,
    TRUE ~ 0)
  if (hitO >= round(XBaSwarmer, 3) ) {
    outsO <- outsO + 1
  }
  #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a runner scored
  if (runnersO >= 4){
    runsO = runsO + 1
    runnersO <-  runnersO - 4
  }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
  if (outsO == 3){
    inningO = inningO + 1
    outsO <- 0
    runnersO <- 0
  }
    
  }

  
}
while(runsC == runsO){
    hitC <- runif(1,0,1)
    #Check if it was a hit or an out, if a hit we add one base runner
    runnersC <- case_when(
      hitC < .171 ~ runnersC + 1,
      hitC < .221 ~ runnersC + 2,
      hitC < .227 ~ runnersC + 3,
      hitC < .256 ~ runnersC + 4,
      TRUE ~ 0)
    if (hitC >= round(XBaLyles,3)) {
      outsC <- outsC + 1
    }
    #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a   runner scored
  
    if (runnersC >= 4){
      runsC = runsC + 1
      runnersC <-  runnersC - 4
    }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
    if (outsC == 3){
      inningC = inningC + 1
      outsC <- 0
      runnersC <- 0
    }
      hitO <- runif(1,0,1)
    #Check if it was a hit or an out, if a hit we add one base runner
    runnersO <- case_when(
      hitO < .148 ~ runnersO + 1,
      hitO < .210 ~ runnersO + 2,
      hitO < .214 ~ runnersO + 3,
      hitO < .248 ~ runnersO + 4,
      TRUE ~ 0)
    if (hitO >= round(XBaSampson,3) ) {
      outsO <- outsO + 1
    }
    #This IF is supposed to add a run if runners get to 4, and returns runners to 3. While adding a run signifying a   runner scored
  
    if (runnersO >= 4){
      runsO = runsO + 1
      runnersO <-  runnersO - 4
    }# This IF checks if there are 3 outs, if there is it resets outs and runners, while changing the inning as well
    if (outsO == 3){
      inningO = inningO + 1
      outsO <- 0
      runnersO <- 0
    }}

finalC[i] <- runsC
finalO[i] <- runsO


}


scores.df$finalC <- finalC
scores.df$finalO <- finalO

scores.df$CubsWin <- ifelse(finalC > finalO, 1, 0)
scores.df$OriolesWin <- ifelse(finalO > finalC, 1, 0)

tableC <- table(scores.df$CubsWin)
tableO <- table(scores.df$OriolesWin)

print(paste0("Predicted Cubs Wins: ", tableC[2], ", predicted Cubs loses: ", tableC[1]))
print(paste0("Predicted Orioles Wins: ", tableO[2], ", predicted Orioles loses: ", tableO[1]))
```

Histograms of Team scores

```{r}
hist(scores.df$finalC, main = "Histogram of Cubs Final Scores", xlab = "Final Scores")
hist(scores.df$finalO, main = "Histogram of Orioles Final Scores", xlab = "Final Scores")


mean(finalC)
mean(finalO)

trying <- scores.df[scores.df$finalC == 2 & scores.df$finalO == 4, ]


```


```{r}

boxplot(scores.df$finalC, xlab = "Cubs", ylab = "Runs Scored")
boxplot(scores.df$finalO, xlab = "Orioles", ylab = "Runs scored")


```







